---
dataMaid: yes
title: MovieLens Project 
subtitle: "HarvardX: PH125.9x Data Science: Capstone:  Sudha Kankipati"
date: 2020-06-10 
output: pdf_document
geometry: margin=2cm
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---

# Introduction
Recommendation systems use rating data from many products and users to make 
recommendations for a specific user.  
Will use the following code to generate datasets and develop algorithm using the edx set.
For a final test of algorithm, We predict movie ratings in the validation set
as if they were unknown.RMSE will be used to evaluate how close predictions are
to the true values in the validation set.

# Data Preparation
 Create edx set, validation set
 Note: this process could take a couple of minutes
```{r}
if(!require(tidyverse))install.packages("tidyverse", repos ="http://cran.us.r-project.org")
if(!require(caret))install.packages("caret", repos ="http://cran.us.r-project.org")
if(!require(data.table))install.packages("data.table", repos ="http://cran.us.r-project.org")
```
 MovieLens 10M dataset:
 https://grouplens.org/datasets/movielens/10m/
 http://files.grouplens.org/datasets/movielens/ml-10m.zip
```{r}
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))


movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")
```
Validation set will be 10% of MovieLens data
```{r}
set.seed(1, sample.kind="Rounding")
```
if using R 3.5 or earlier, use `set.seed(1)` instead
```{r}
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]
```
Make sure userId and movieId in validation set are also in edx set
```{r}
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")
```
 Add rows removed from validation set back into edx set
```{r}
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```
# Data Exploration
Before creating the model,Understanding the features of the rating data set.
This step will help build a better model.

```{r , echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library("ggplot2")
library("pander")
```

```{r}
ggAggHist <- getFromNamespace("ggAggHist", "dataMaid")
ggAggBarplot <- getFromNamespace("ggAggBarplot", "dataMaid")
```

## Data report overview
The dataset examined has the following dimensions:

----------------------------------
Feature                     Result
------------------------ ---------
Number of observations     9000055

Number of variables              6
----------------------------------


### Checks performed
The following variable checks were performed, depending on the data type of each variable:

---------------------------------------------------------------------------------------------------------------------------------------------------
&nbsp;                                                 character    factor    labelled   haven labelled   numeric    integer    logical     Date   
----------------------------------------------------- ----------- ---------- ---------- ---------------- ---------- ---------- --------- ----------
Identify miscoded missing values                       $\times$    $\times$   $\times$      $\times$      $\times$   $\times$             $\times$ 

Identify prefixed and suffixed whitespace              $\times$    $\times$   $\times$      $\times$                                               

Identify levels with < 6 obs.                          $\times$    $\times$   $\times$      $\times$                                               

Identify case issues                                   $\times$    $\times$   $\times$      $\times$                                               

Identify misclassified numeric or integer variables    $\times$    $\times$   $\times$      $\times$                                               

Identify outliers                                                                                         $\times$   $\times$             $\times$ 
---------------------------------------------------------------------------------------------------------------------------------------------------



Please note that all numerical values in the following have been rounded to 2 decimals.


### Summary table

---------------------------------------------------------------------------------------
&nbsp;        Variable class     # unique values   Missing observations  Any problems? 
------------- ---------------- ----------------- ---------------------- ---------------
[userId]      integer                      69878                 0.00 %                

[movieId]     numeric                      10677                 0.00 %    $\times$    

[rating]      numeric                         10                 0.00 %    $\times$    

[timestamp]   integer                    6519590                 0.00 %                

[title]       character                        1               100.00 %    $\times$    

[genres]      character                        1               100.00 %    $\times$    
---------------------------------------------------------------------------------------




## Variable list
### userId

\bminione

----------------------------------------
Feature                           Result
------------------------- --------------
Variable type                    integer

Number of missing obs.           0 (0 %)

Number of unique values            69878

Median                             35738

1st and 3rd quartiles       18124; 53607

Min. and max.                   1; 71567
----------------------------------------


\emini
\bminitwo
```{r 'Var-1-userId', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 5000, 10000, 15000, 
20000, 25000, 30000, 35000, 40000, 45000, 50000, 55000, 60000, 
65000, 70000), xmax = c(5000, 10000, 15000, 20000, 25000, 30000, 
35000, 40000, 45000, 50000, 55000, 60000, 65000, 70000, 75000
), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(586293L, 
643123L, 624851L, 643486L, 636445L, 640125L, 631925L, 632813L, 
631573L, 623177L, 617584L, 617321L, 632097L, 629695L, 209547L
)), class = "data.frame", row.names = c(NA, -15L)), vnam = "userId")
```

\emini




\fullline

### movieId

\bminione

-------------------------------------
Feature                        Result
------------------------- -----------
Variable type                 numeric

Number of missing obs.        0 (0 %)

Number of unique values         10677

Median                           1834

1st and 3rd quartiles       648; 3626

Min. and max.                1; 65133
-------------------------------------


\emini
\bminitwo
```{r 'Var-2-movieId', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 2000, 4000, 6000, 
8000, 10000, 12000, 14000, 16000, 18000, 20000, 22000, 24000, 
26000, 28000, 30000, 32000, 34000, 36000, 38000, 40000, 42000, 
44000, 46000, 48000, 50000, 52000, 54000, 56000, 58000, 60000, 
62000, 64000), xmax = c(2000, 4000, 6000, 8000, 10000, 12000, 
14000, 16000, 18000, 20000, 22000, 24000, 26000, 28000, 30000, 
32000, 34000, 36000, 38000, 40000, 42000, 44000, 46000, 48000, 
50000, 52000, 54000, 56000, 58000, 60000, 62000, 64000, 66000
), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 
0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(4729781L, 
2327700L, 920786L, 447994L, 162822L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 4394L, 36334L, 0L, 34167L, 43208L, 22262L, 16798L, 13628L, 
23194L, 10991L, 37955L, 21373L, 35031L, 18137L, 21590L, 29884L, 
15466L, 15251L, 8165L, 2504L, 640L)), class = "data.frame", row.names = c(NA, 
-33L)), vnam = "movieId")
```

\emini


- Note that the following possible outlier values were detected: \"25736\", \"25737\", \"25744\", \"25746\", \"25750\", \"25752\", \"25753\", \"25755\", \"25757\", \"25759\" (2351 additional values omitted). 



\fullline

### rating

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          10

Median                            4

1st and 3rd quartiles          3; 4

Min. and max.                0.5; 5
-----------------------------------


\emini
\bminitwo
```{r 'Var-3-rating', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0.4, 0.6, 0.8, 1, 1.2, 
1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8, 
4, 4.2, 4.4, 4.6, 4.8), xmax = c(0.6, 0.8, 1, 1.2, 1.4, 1.6, 
1.8, 2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8, 4, 4.2, 4.4, 
4.6, 4.8, 5), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 
0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(85374L, 0L, 345679L, 
0L, 0L, 106426L, 0L, 711422L, 0L, 0L, 333010L, 0L, 2121240L, 
0L, 0L, 791624L, 0L, 2588430L, 0L, 0L, 526736L, 0L, 1390114L)), class = "data.frame", row.names = c(NA, 
-23L)), vnam = "rating")
```

\emini


- Note that the following possible outlier values were detected: \"4.5\", \"5\". 



\fullline

### timestamp

\bminione

-------------------------------------------------
Feature                                    Result
------------------------- -----------------------
Variable type                             integer

Number of missing obs.                    0 (0 %)

Number of unique values                   6519590

Median                                 1035493918

1st and 3rd quartiles       946768283; 1126750881

Min. and max.               789652009; 1231131736
-------------------------------------------------


\emini
\bminitwo
```{r 'Var-4-timestamp', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(780000000L, 800000000L, 
820000000L, 840000000L, 860000000L, 880000000L, 900000000L, 920000000L, 
940000000L, 960000000L, 980000000L, 1000000000L, 1020000000L, 
1040000000L, 1060000000L, 1080000000L, 1100000000L, 1120000000L, 
1140000000L, 1160000000L, 1180000000L, 1200000000L, 1220000000L
), xmax = c(800000000L, 820000000L, 840000000L, 860000000L, 880000000L, 
900000000L, 920000000L, 940000000L, 960000000L, 980000000L, 1000000000L, 
1020000000L, 1040000000L, 1060000000L, 1080000000L, 1100000000L, 
1120000000L, 1140000000L, 1160000000L, 1180000000L, 1200000000L, 
1220000000L, 1240000000L), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 
0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(2L, 0L, 470532L, 
639530L, 227202L, 82809L, 141303L, 192846L, 928486L, 765087L, 
444427L, 378206L, 310697L, 415366L, 408194L, 417369L, 761073L, 
488977L, 422164L, 445494L, 370177L, 386887L, 303227L)), class = "data.frame", row.names = c(NA, 
-23L)), vnam = "timestamp")
```

\emini




\fullline

### title

* The variable only takes one value: \"NA\".



\fullline

### genres

* The variable only takes one value: \"NA\".



\fullline

### showing train set
```{r}
edx %>% as_tibble()

edx %>% summarize(
  n_users=n_distinct(userId),# unique users from train set
  n_movies=n_distinct(movieId),# unique movies from train set
  min_rating=min(rating), # the lowest rating in train set
  max_rating=max(rating) # the highest rating in train set
)
```

\fullline
# Data Cleaning
## Removing genres and timestamp.
```{r}
edx <- edx %>% select(userId, movieId, rating, title)
validation <- validation %>% select(userId, movieId, rating, title)
```

\fullline
#  Data Visualization
## Movies exploration
Checking number of different movies present in 'edx' set
```{r}
length(unique(edx$movieId))
```
Distribution of movies: Movies rated more than others (histogram)
```{r , warning=FALSE, message=FALSE, error=FALSE}
edx %>% group_by(movieId) %>%
  summarise(n=n()) %>%
  ggplot(aes(n)) +
  geom_histogram(bin = 30, color = "white") +
  scale_x_log10() + 
  ggtitle("Movies")
```

## Users exploration
Checking number of different users present in 'edx' set
```{r}
length(unique(edx$userId))
```
Distribution of users
```{r, warning=FALSE, message=FALSE, error=FALSE}
edx %>% group_by(userId) %>%
  summarise(n=n()) %>%
  arrange(n) %>%
  head()
```
\fullline

## Insights
* The first thing we notice is that some movies get rated more than others.
Here is the distribution: plot count rating by movie
```{r}
edx %>%
  count(movieId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 30, color = "White") +
  scale_x_log10() +
  ggtitle("Movies")
```
* Our second observation is that some users are more active than others at rating movies:
plot count rating by user
```{r}
edx %>%
  count(userId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 30, color = "green") +
  scale_x_log10() +
  ggtitle("Users")
```

\fullline  

# Methods and Analysis

## RMSE
Root Mean Squared Error (RMSE) is the indicator used to compare the predicted value 
with the actual outcome. During the model development, we use the test(edx) set to 
predict the outcome. 
When the model is ready, then we use the 'validation' set.

### Linear Model
Average movie rating model 

* compute mean rating
```{r}
mu <- mean(edx$rating) 
mu
```
* compute root mean standard error
```{r}
naive_rmse <- RMSE(edx$rating, mu) 
```

* Check results and  Save prediction in tibble
```{r}
rmse_results <- tibble(method = "Average movie rating model", RMSE = naive_rmse)
rmse_results %>% knitr::kable()
```
### Movie effect model

* Simple model taking into account the movie effect b_i
Subtract the rating minus the mean for each rating the movie received
Plot number of movies with the computed b_i
```{r , warning=FALSE, message=FALSE, error=FALSE}
movie_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - mu))
movie_avgs %>% qplot(b_i, geom ="histogram", bins = 10, data = ., color = I("White"),
         ylab = "Number of movies", main = "Number of movies with the computed b_i")
```
* Test and save rmse results 
```{r, warning=FALSE, message=FALSE, error=FALSE}
predicted_ratings <- mu +  validation %>%
  left_join(movie_avgs, by='movieId') %>%
  pull(b_i)
model_1_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results,
                          tibble(method="Movie effect model",  
                                     RMSE = model_1_rmse ))
```
* Check results
```{r}
rmse_results %>% knitr::kable()
```

## Matrix factorization
recosystem is a package for recommendation system
using Matrix Factorization. High performance multi-core
parallel computing is supported in this package.

```{r}
if(!require(recosystem)) 
  install.packages("recosystem", repos = "http://cran.us.r-project.org")
set.seed(1)
```
### This is a randomized algorithm

### Converting the train and test sets into recosystem input format
```{r, warning=FALSE, message=FALSE, error=FALSE}
train_data <-  with(edx, data_memory(user_index = userId, 
                                           item_index = movieId, 
                                           rating     = rating))
test_data  <-  with(validation,  data_memory(user_index = userId, 
                                           item_index = movieId, 
                                           rating     = rating))
```
### Create the model object
```{r, warning=FALSE, message=FALSE, error=FALSE}
recommender <-  recosystem::Reco()
```

### Select the best tuning parameters
```{r, warning=FALSE, message=FALSE, error=FALSE}
opts <- recommender$tune(train_data, opts = list(dim = c(10, 20, 30), 
                                       lrate = c(0.1, 0.2),
                                       costp_l2 = c(0.01, 0.1), 
                                       costq_l2 = c(0.01, 0.1),
                                       nthread  = 4, niter = 10))
```

### Training the algorithm  
```{r}
recommender$train(train_data, opts = c(opts$min, nthread = 4, niter = 20))
```

### Calculate the predicted values  

* We predict unknown entries in the rating matrix on test set.
use the `$predict()` method to compute predicted values
return predicted values in memory
```{r}
predicted_ratings <-  recommender$predict(test_data, out_memory())
head(predicted_ratings, 10)#We use the test set for the final assessment
```

* ceiling rating at 5
```{r}
ind <- which(predicted_ratings > 5)
predicted_ratings[ind] <- 5
```

* floor rating at 0.50
```{r}
ind <- which(predicted_ratings < 0.5)
predicted_ratings[ind] <- 0.5
```
* create a results table with this approach
```{r}
model_2_rmse <- RMSE(validation$rating, predicted_ratings)
rmse_results <- bind_rows(rmse_results,
                          tibble(method="Movie and User + Matrix Fact. on test set",
                                 RMSE = model_2_rmse))
rmse_results %>% knitr::kable()
set.seed(1)
```

### Converting 'edx' and 'validation' sets to recosystem input format
```{r}
edx_reco <-  with(edx, data_memory(user_index = userId, 
                                   item_index = movieId, 
                                   rating = rating))
validation_reco  <-  with(validation, data_memory(user_index = userId, 
                                                  item_index = movieId, 
                                                  rating = rating))
```

### Create the model object
```{r}
recommender <-  recosystem::Reco()
```
### Tune the parameters
```{r}
opts <-  recommender$tune(edx_reco, opts = list(dim = c(10, 20, 30), 
                                      lrate = c(0.1, 0.2),
                                      costp_l2 = c(0.01, 0.1), 
                                      costq_l2 = c(0.01, 0.1),
                                      nthread  = 4, niter = 10))
```

# Training the model
```{r}
recommender$train(edx_reco, opts = c(opts$min, nthread = 4, niter = 20))
```
# Results
* create a results table with this and prior approaches
Calculate the prediction
```{r}
predicted_ratings <-  recommender$predict(validation_reco, out_memory())


model_3_rmse <- RMSE(validation$rating, predicted_ratings)
rmse_results <- bind_rows(rmse_results,
                          tibble(method="Movie and User effects and Matrix Fact. on validation set",
                                 RMSE = model_3_rmse))
rmse_results %>% knitr::kable()
```

# Conclusion   
We reach a RMSE of `r model_3_rmse ` on validation set using the full edx set for training. 
The model Movie and User effects and Matrix Factorization achieved this performance using the
methodology presented in the course’s book 
(https://rafalab.github.io/dsbook/largedatasets.html#recommendation-systems)

Report generation information:
 *  Created by: Sudha Kankipati
 *  Report creation time: Wed Jun 10 2020 
 *  R version 4.0.0 (2020-04-24).
 *  Platform: x86_64-w64-mingw32/x64 (64-bit)(Windows 10 x64 (build 18363)).
 * https://github.com/DrSudhaK/MovieLensProject.git

 
